---
title: "Case Study 02"
author: "Hien Lam"
date: "11/29/2021"
output: html_notebook
abstract: "How does company culture affect attrition?"
---
## Read and tidy the data

Load the necessary packages
```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(viridis)
library(ggridges)
library(cowplot)
library(lemon) # uninstall if not used
library(caret) #confusion matrix
library(e1071)
library(class) #knn.cv
```
Read the data
```{r message=FALSE, warning=FALSE}
attr <- read_csv('https://raw.githubusercontent.com/hmlam1/CaseStudy02DDS/main/CaseStudy02.csv', show_col_types=F)
```
Study the data: 870 rows (observations) and 36 columns (variables). First 10 rows are printed to allow reader to click through the columns.
```{r}
dim(attr)
spec(attr)
head(attr, n=10)
```
Remove columns with single unique value as they are not useful for EDA or prediction
```{r}
attr <- attr %>% select(-c('Over18', 'EmployeeCount', 'EmployeeNumber', 'PerformanceRating', 'StandardHours'))
```

Confirm that there are zero NA values
```{r}
sum(as.numeric(is.na.data.frame(attr)))
```
I copied *Attrition* into a new column called *attr10*, replaced "No" with 0 and "Yes" with 1, set class as numeric for aggregate function to calculate attrition percentages.
```{r}
attr <- attr %>% mutate(attr10 = Attrition, attr10 = replace(attr10, attr10 != "No", 1))
attr <- attr %>% mutate(attr10 = replace(attr10, attr10 != "1", 0))
attr$attr10 <- as.numeric(attr$attr10)
```

---

## Study the data

Observations:

Age

- Distribution is right skewed; must log transform prior to modeling  
- Age vs attrition is very significant; highest between 18-22 at >50%
- Age vs job role is significant  
- Age vs gender and department is about even
```{r}
# dist of age
hist(attr$Age, breaks=50, main='Distribution of Age', xlab = 'Age')
quantile(attr$Age)
# vs gender, department, job role
boxplot(Age~Gender, data=attr) # insig
boxplot(Age~Department, data=attr) # significant ish
boxplot(Age~JobRole, data=attr) # significant
# vs edu?
# edu vs gender?
```
Attrition (as it relates to personal self/life)

- 140 attrition, 730 no attrition  
- vs gender is about even (15% female, 17% male)  
- Age vs attrition is very significant; highest between 18-22 at >50%  
- Highest for single marital status at 26%  
- Increases as distance from home goes up; 15%+ for 12.2+ miles  
- Increases as education level goes down  
- Highest for hr 27%, tech degree and marketing 20-23% (edu field)  
- Highest for sales department at 22%   
- Highest for sales rep at 45%, hr/lab tech 20-22% (job role)  
- Highest for lowest job level at 26%  
- Highest for 5> total working years at 31%  
- Highest for 5+ number of companies worked all >20%

Attrition (as it relates to work culture)

- Highest for lowest perceived level of job involvement at 47%  
- Highest for lowest perceived level of work life balance at 35%  
- Highest for overtime at 32%  
- Attr increases with decreasing years in current role; 0-3 years 30%  
- Highest for the least (26%) and best (22%) stock option level. Take a closer look  
- Highest for lowest perceived level of environment satisfaction at 24%  
- Highest for lowest perceived level of job satisfaction at 21%  
- Highest for the first 3 years at company. Will exclude this var as it will skew the model bc not all the years have equal representation
- Highest for travel frequently for business travel at 22%  
- Highest for first 2 years in current role at 23%  
- Highest for lowest relationship satisfaction with manager at 20%  
- Higher for increasing years since last promotion. Uneven distribution. Probably should not include this in the model  
- Highest for initial year with current manager however it is skewed for 14 year mark. Either remove that value or don't include variable in model
- Inconclusive for training times last year
- Inconclusive for percent salary hike. Need to look closer against other variables
```{r eval=FALSE, include=FALSE}
# attr count. 0 = "No" 1 = "Yes"
attr %>% count(Attrition)
hist(attr$attr10)
# vs gender. insig
aggregate(attr10~Gender, data=attr, mean)
# vs age. sig
attr.age <- aggregate(x = attr$attr10, by = list(cut(attr$Age, 10)), mean)
names(attr.age) <- c('Age', 'Attrition')
ggplot(attr.age, aes(x = Age, y = Attrition)) + geom_bar(stat = "identity")
# vs marital status. sig
attr.mar <- aggregate(attr10~MaritalStatus, data=attr, mean)
ggplot(attr.mar, aes(x = MaritalStatus, y = attr10)) + geom_bar(stat = "identity")
# vs distance from home. sig
attr.dist <- aggregate(x = attr$attr10, by = list(cut(attr$DistanceFromHome, 10)), mean)
names(attr.dist) <- c('dist', 'Attrition')
ggplot(attr.dist, aes(x = dist, y = Attrition)) + geom_bar(stat = "identity")
# education level. sig if comparing lowest and highest edu levels
attr.edu <- aggregate(attr10~Education, data=attr, mean)
ggplot(attr.edu, aes(x = Education, y = attr10)) + geom_bar(stat = "identity")
# education field. sig
attr.eduf <- aggregate(attr10~EducationField, data=attr, mean)
ggplot(attr.eduf, aes(x = EducationField, y = attr10)) + geom_bar(stat = "identity")
# total working years. very sig
attr.work <- aggregate(x = attr$attr10, by = list(cut(attr$TotalWorkingYears, 8)), mean)
names(attr.work) <- c('total', 'Attrition')
ggplot(attr.work, aes(x = total, y = Attrition)) + geom_bar(stat = "identity")
# number of companies worked. sig
attr.ncomp <- aggregate(attr10~NumCompaniesWorked, data=attr, mean)
ggplot(attr.ncomp, aes(x = NumCompaniesWorked, y = attr10)) + geom_bar(stat = "identity")
# vs department. sig
attr.dep <- aggregate(attr10~Department, data=attr, mean) 
ggplot(attr.dep, aes(x = Department, y = attr10)) + geom_bar(stat = "identity")
# vs job role. very sig
attr.role <- aggregate(attr10~JobRole, data=attr, mean) 
ggplot(attr.role, aes(x = JobRole, y = attr10)) + geom_bar(stat = "identity")
# job level. sig
attr.level <- aggregate(attr10~JobLevel, data=attr, mean)
ggplot(attr.level, aes(x = JobLevel, y = attr10)) + geom_bar(stat = "identity")
```

```{r eval=FALSE, include=FALSE}
# job involvement. very sig
attr.inv <- aggregate(attr10~JobInvolvement, data=attr, mean)
ggplot(attr.inv, aes(x = JobInvolvement, y = attr10)) + geom_bar(stat = "identity")
# job satisfaction. sig
attr.satis <- aggregate(attr10~JobSatisfaction, data=attr, mean)
ggplot(attr.satis, aes(x = JobSatisfaction, y = attr10)) + geom_bar(stat = "identity")
# vs work life balance. very sig
attr.wlb <- aggregate(attr10~WorkLifeBalance, data=attr, mean)
ggplot(attr.wlb, aes(x = WorkLifeBalance, y = attr10)) + geom_bar(stat = "identity")
# years at company. sig but misleading
attr.yrcomp <- aggregate(x = attr$attr10, by = list(cut(attr$YearsAtCompany, 20)), mean)
names(attr.yrcomp) <- c('Years', 'Attrition')
ggplot(attr.yrcomp, aes(x = Years, y = Attrition)) + geom_bar(stat = "identity")
# years in current role
attr.yrcur <- aggregate(x = attr$attr10, by = list(cut(attr$YearsInCurrentRole, 9)), mean)
names(attr.yrcur) <- c('Years', 'Attrition')
ggplot(attr.yrcur, aes(x = Years, y = Attrition)) + geom_bar(stat = "identity")
# vs business travel. sig
attr.travel <- aggregate(attr10~BusinessTravel, data=attr, mean)
ggplot(attr.travel, aes(x = BusinessTravel, y = attr10)) + geom_bar(stat = "identity")
# vs environment satisfaction. sig
attr.env <- aggregate(attr10~EnvironmentSatisfaction, data=attr, mean)
ggplot(attr.env, aes(x = EnvironmentSatisfaction, y = attr10)) + geom_bar(stat = "identity")
# vs overtime. very sig
attr.ot <- aggregate(attr10~OverTime, data=attr, mean)
ggplot(attr.ot, aes(x = OverTime, y = attr10)) + geom_bar(stat = "identity")
# vs stock option level.
attr.stock <- aggregate(attr10~StockOptionLevel, data=attr, mean)
ggplot(attr.stock, aes(x = StockOptionLevel, y = attr10)) + geom_bar(stat = "identity")
# vs percent salary hike. insig. might be related to a diff variable
attr.inc <- aggregate(attr10~PercentSalaryHike, data=attr, mean)
ggplot(attr.inc, aes(x = PercentSalaryHike, y = attr10)) + geom_bar(stat = "identity")
# vs years since last promotion. sig but inconsistent
attr.yrpromo <- aggregate(attr10~YearsSinceLastPromotion, data=attr, mean)
ggplot(attr.yrpromo, aes(x = YearsSinceLastPromotion, y = attr10)) + geom_bar(stat = "identity")
# vs training times last year. sig but inconsistent
attr.train <- aggregate(attr10~TrainingTimesLastYear, data=attr, mean)
ggplot(attr.train, aes(x = TrainingTimesLastYear, y = attr10)) + geom_bar(stat = "identity")
# vs years with current manager. sig but inconsistent
attr.yrman <- aggregate(attr10~YearsWithCurrManager, data=attr, mean)
ggplot(attr.yrman, aes(x = YearsWithCurrManager, y = attr10)) + geom_bar(stat = "identity")
# vs relationship satisfaction w manager. barely sig
attr.mansatis <- aggregate(attr10~RelationshipSatisfaction, data=attr, mean)
ggplot(attr.mansatis, aes(x = RelationshipSatisfaction, y = attr10)) + geom_bar(stat = "identity")
# Rate: hourly, daily, monthly, monthly income. Plot one
```

Miscellaneous ways to check correlation. ignore for now
``` {r}
attr.quant <- select_if(attr, is.numeric)
plot(attr[,c(2:10)], col = 'royalblue')
colnames(attr.quant)
symnum(cor(attr.quant[,c(2:24)], use='complete.obs'))
```

---

## EDA

### Years in current role vs attr

30% attr in first year. 23% in first 3 years.
```{r}
# density of YearsInCurrentRole with attr == yes overlay. WHY does it look weird?!
yrs.attr.yes <- attr %>% select(YearsInCurrentRole, Attrition) %>% filter(Attrition == 'Yes')

ggplot() +
geom_density(data = attr, aes(x = YearsInCurrentRole),
fill = "lightgray", color = "grey", alpha = 0.85) +
geom_density(data = yrs.attr.yes, aes(x = YearsInCurrentRole),
fill = "royalblue", color = "royalblue", alpha = 0.2)

# density plot of YearsinCurrentRole with attr
attr %>% ggplot(aes(YearsInCurrentRole, fill=Attrition)) + geom_density(alpha=0.5)

# count instead of density
ggplot(attr, aes(YearsInCurrentRole, y=..count.., fill=Attrition)) +
geom_density(color = "black", alpha = 0.7) +
scale_fill_manual(values = c("#E69F00", "#56B4E9"))

ggplot(attr, aes(YearsInCurrentRole, y=..count..)) +
geom_density(color = "black", alpha = 0.7, fill='royalblue')
```
### Job involvement vs attr
```{r}
# density plot
ggplot(data=attr, aes(x=Attrition, group=JobInvolvement, fill=JobInvolvement)) +
geom_density(adjust=1.25, position="fill")

# heatmap. change 1-4 to very unsatisfied - very satisfied?
attr.ji <- attr %>% count(Attrition, JobInvolvement)
attr.ji <- pivot_wider(attr.ji, names_from = Attrition, values_from = n)
attr.ji <- attr.ji %>% mutate(sum = rowSums(attr.ji[,c(2:3)]))
attr.ji <- pivot_longer(attr.ji, c(2:3), names_to='Attrition', values_to = "value")
attr.ji <- attr.ji %>% mutate(perc=value/sum*100)
# plot without borders
ggplot(attr.ji, aes(x=JobInvolvement, y=Attrition, fill=perc)) + geom_tile() + geom_text(aes(label = perc), color = "white", size = 4) + coord_fixed() + scale_fill_viridis(option='F', begin=.2, direction=-1) + guides(fill = guide_colourbar(title='Attrition \nPercent', ticks = FALSE, barwidth = .75, barheight = 20))
# with borders
ggplot(attr.ji, aes(x=JobInvolvement, y=Attrition, fill=perc)) + geom_tile(color = "white", lwd = 1.5, linetype = 1) + geom_text(aes(label = perc), color = "white", size = 4) + coord_fixed() + scale_fill_viridis(option='F', begin=.2, direction=-1) + guides(fill = guide_colourbar(title='Attrition \nPercent', ticks = FALSE, barwidth = .75, barheight = 20))
```
### Job role vs monthly income
```{r}
ggplot(attr, aes(x = MonthlyIncome, y = JobRole, fill = stat(x))) +
geom_density_ridges_gradient(scale = 2.5, rel_min_height = 0.01) +
scale_fill_viridis(name = "Monthly Income", option = "F", direction=-1, begin=.15, end=.9)

# job role vs attrition. sales rep 45% hr 22%
# unsurprising, income drives attrition but is there an underlying cause?
role.attr <-  attr %>% count(Attrition, JobRole)
role.attr <- pivot_wider(role.attr, names_from = Attrition, values_from = n)
role.attr <- role.attr %>% mutate(Perc = Yes/(No+Yes))
arrange(role.attr, -Perc)

# fail. job role vs work life balance
role.wlb <-  attr %>% count(WorkLifeBalance, JobRole)
role.wlb <- pivot_wider(role.wlb, names_from = WorkLifeBalance, values_from = n)
role.wlb %>% mutate(perc = rowMeans(role.wlb[,c(2:5)], na.rm=TRUE))

# job role vs job satisfaction
## replace 1-4 with very unsatisfied-very satisfied
attr$JobSatisfaction <- ifelse(attr$JobSatisfaction == 1, 'Very unsatisfied', attr$JobSatisfaction)
attr$JobSatisfaction <- ifelse(attr$JobSatisfaction == 2, 'Unsatisfied', attr$JobSatisfaction)
attr$JobSatisfaction <- ifelse(attr$JobSatisfaction == 3, 'Satisfied', attr$JobSatisfaction)
attr$JobSatisfaction <- ifelse(attr$JobSatisfaction == 4, 'Very satisfied', attr$JobSatisfaction)
## % of job satis level 1 for each job role 
role.satis <-  attr %>% count(JobSatisfaction, JobRole)
role.satis <- pivot_wider(role.satis, names_from = JobSatisfaction, values_from = n)
role.satis <- role.satis %>% mutate(sum = rowSums(role.satis[,c(2:5)], na.rm=TRUE), perc=`1`/sum)
arrange(role.satis, -perc)
# % of env satis level 1 for each job role
role.env <-  attr %>% count(EnvironmentSatisfaction, JobRole)
role.env <- pivot_wider(role.env, names_from = EnvironmentSatisfaction, values_from = n)
role.env <- role.env %>% mutate(sum = rowSums(role.env[,c(2:5)], na.rm=TRUE), perc=`1`/sum)
arrange(role.env, -perc)
```

---

## Analysis (backed up with robust experimentation & appropriate visualizations)

### Identify top three factors that contribute to turnover

1. Years in current role  
2. Job involvement  
3. Work life balance
```{r}

```
### Job role specific trends

Display monthly salary for each role and their attrition %. Attr is high for lower salary. Nothing new here. Let's dive deeper.
```{r}

```
### Other interesting trends or observations

Unsurprising, income drives attrition but is there an underlying cause? Looked at wlb, job & env satisfaction. sales rep is high (favorable) for those factors yet high attr so income matters to a threshold then attr decreases when income is high yet low cultural variables.
```{r}

```

---

## Prediction using KNN, Naïve Bayes, and Linear Regression 

### Attrition (min. 60% sensi & speci)
```{r}
#### KNN -- Not Running
set.seed(6)
splitPerc = .70
summary(attr.quant)
trainIndices = sample(1:dim(attr.quant)[1],round(splitPerc * dim(attr.quant)[1]))
trainIndices
train = attr.quant[trainIndices,]
test = attr.quant[-trainIndices,]

train


# k = 3
classifications = knn(train[,c(1,4)],test[,c(1,4)],train$Attrition, prob = TRUE, k = 3)
table(classifications,test$Species)
confusionMatrix(table(classifications,test$Species))
# k = 5
classifications = knn(train[,c(1,2)],test[,c(1,2)],train$Attrition, prob = TRUE, k = 5)
table(test$Species,classifications)
confusionMatrix(table(test$Species,classifications))
# k = 10
classifications = knn(train[,c(1,2)],test[,c(1,2)],train$Attrition, prob = TRUE, k = 10)
table(test$Species,classifications)
confusionMatrix(table(test$Species,classifications))
# k = 20
classifications = knn(train[,c(1,2)],test[,c(1,2)],train$Attrition, prob = TRUE, k = 20)
table(test$Species,classifications)
CM = confusionMatrix(table(test$Species,classifications))
CM$overall[1]


#### Naïve Bayes #works :)
model = naiveBayes(Attrition~.,data = attr)
predict(model,attr[,c(16,17)])
predict(model,attr[,c(1,2,4,5,6,7,8,9,10,11,12,13,14,15,16,17,18,19,20,21,22,23,24,25,26,27,28,29,30,31,32)])
df = data.frame(MaritialStatus = "Single", JobSatisfaction = "Very unsatisfied")
predict(model,df) #just classifications
predict(model,df, type = "raw") #gives probabilities 
#           No       Yes
#[1,] 0.7877095 0.2122905


#### Linear Regression -- Finding best model 
# need to utilize numeric values only

#degree 1 model - Age vs Attrition
fit = lm(Age~attr10, attr.quant)
summary(fit)
confint(fit)

#degree 1 model - Daily Rate vs Attrition
fit = lm(DailyRate~attr10, attr.quant)
summary(fit)
confint(fit)

#degree 1 model - DistanceFromHome vs Attrition
fit = lm(DistanceFromHome~attr10, attr.quant)
summary(fit)
confint(fit)

#degree 1 model - Environment vs Attrition
fit = lm(EnvironmentSatisfaction~attr10, attr.quant)
summary(fit)
confint(fit)

#degree 1 model - Education vs Attrition
fit = lm(Education~attr10, attr.quant)
summary(fit)
confint(fit)

#degree 1 model - HourlyRate vs Attrition
fit = lm(HourlyRate~attr10, attr.quant)
summary(fit)
confint(fit)

#degree 1 model - JobInvolvenemtn vs Attrition
fit = lm(JobInvolvement~attr10, attr.quant)
summary(fit)
confint(fit)

#degree 1 model - JobLevel vs Attrition
fit = lm(JobLevel~attr10, attr.quant)
summary(fit)
confint(fit)

#degree 1 model - JobSatisfaction vs Attrition
fit = lm(JobSatisfaction~attr10, attr.quant)
summary(fit)
confint(fit)

#degree 1 model - MonthlyIncome vs Attrition
fit = lm(MonthlyIncome~attr10, attr.quant)
summary(fit)
confint(fit)

#degree 1 model - MonthlyRate vs Attrition
fit = lm(MonthlyRate~attr10, attr.quant)
summary(fit)
confint(fit)

#degree 1 model - NumCompaniesWorked vs Attrition
fit = lm(NumCompaniesWorked~attr10, attr.quant)
summary(fit)
confint(fit)

#degree 1 model - PercentSalarayHike vs Attrition
fit = lm(PercentSalaryHike~attr10, attr.quant)
summary(fit)
confint(fit)

#degree 1 model - RelationshipSatisfaction vs Attrition
fit = lm(RelationshipSatisfaction~attr10, attr.quant)
summary(fit)
confint(fit)

#degree 1 model - StockOptionLevel vs Attrition
fit = lm(StockOptionLevel~attr10, attr.quant)
summary(fit)
confint(fit)

#degree 1 model - TotalWorkingYears vs Attrition
fit = lm(TotalWorkingYears~attr10, attr.quant)
summary(fit)
confint(fit)

#degree 1 model - TrainingTimesLastYear vs Attrition
fit = lm(TrainingTimesLastYear~attr10, attr.quant)
summary(fit)
confint(fit)

#degree 1 model - Worklifebalance vs Attrition
fit = lm(WorkLifeBalance~attr10, attr.quant)
summary(fit)
confint(fit)

#degree 1 model - YEarsAtCompanye vs Attrition
fit = lm(YearsAtCompany~attr10, attr.quant)
summary(fit)
confint(fit)

#degree 1 model - YearsInCurrentRole vs Attrition
fit = lm(YearsInCurrentRole~attr10, attr.quant)
summary(fit)
confint(fit)

#degree 1 model - YearsSinceLastPromotion vs Attrition
fit = lm(YearsSinceLastPromotion~attr10, attr.quant)
summary(fit)
confint(fit)

#degree 1 model - YearsWithCurrentManager vs Attrition
fit = lm(YearsWithCurrManager~attr10, attr.quant)
summary(fit)
confint(fit)





#degree 2 model - if necessary
mtcars %>% ggplot(aes(x = wt, y = mpg)) + geom_point()
mtcars3 = mtcars %>% mutate(wt2 = wt^2)
fit = lm(mpg~wt+wt2, mtcars3)
summary(fit)




#

```
### Monthly salary (RMSE <3000)
```{r}

```

## Appendix

Plots from observations that were not shown for brevity
